---
output:
  html_document: default
  pdf_document: default
---

```{r, echo=FALSE}
library(knitr)
#opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

---
title: "PCA: genetic simulation and wonkiness "
author: "Trevor Faske"
output: pdf_document
---

This Rmarkdown is a simulates genetic data and highlights different issues with how PCA (or most clustering methods) partitions variance and clusters data. Following many examples from dumb paper: Elhaik et al. 2022 *Scientific Reports*: [<https://doi.org/10.1038/s41598-022-14395-4>]

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Desktop/projects/mixed_ploidy/')
```

### Loading packages

```{r eval=TRUE, eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE}
library(tidyverse)
library(ggsci)
library('wesanderson')
library(scales)
library(patchwork)
library(ggpubr)
```

## Simulate data function

creates ancestral allele frequencies from rbeta() then creates new populations and samples individual genotypes using rbinom() assuming all diploid.

#### features:

-   **nloci:** number of loci\
-   **nind:** number of individuals total\
-   **npop:** number of populations\
-   **theta1:** shape1 of beta distribution (could be thought of as theta diversity)\
-   **theta2:** shape2 of beta distribution (could be thought of as theta diversity)
    -   note: use hist(rbeta(1000,theta1,theta2)) to view allele frequency distribution\
-   **Fst:** population differentiation based on a Wright-Fisher model\
-   **MAF:** Minor allele frequeuncy cutoff\
-   **sampleSize:** sample size proportions options for populations
    -   'equal' = equal population sizes (ind per pop = nind/npop)\
    -   'unequal' = sample sizes vary among individuals within a population sample sizes vary (see popProb for sampling)\
-   **popProb:** sample probabilities of each population (length must be equal to npop and sum to 1). If NULL, popProb will be equal to 1/npop for each population. (ex. npop = 5, popProb = c(.2,.2,.2,.2,.2))\
-   **maxMissing:** max amount of mising data per individual (sampled using a scaled rpois())\
-   **seed:** seed to be set for random sampling

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=5,fig.height=6}


sim_mixedploidy <- function(nloci=5000,nind=300,npop=10,theta1=0.8,theta2=0.8,
                            Fst=.1,MAF=0.05,sampleSize='equal',
                            popProb=NULL,maxMissing=NULL,seed=666){

  set.seed(seed)

  # ancestral population
  anc.pi<-rbeta(nloci, theta1, theta2)

  # generate population allele freqs, double nloci to remove MAF and invariants later
  sim.pop <- matrix(nrow=nloci*2, ncol=npop)
  for(j in 1:npop){
    Fm <- (1-Fst)/Fst #equal to the F-model. Nicholson et al 2002
    sim.pop[,j] <- rbeta(nloci*2, anc.pi * Fm,
                         (1-anc.pi) * Fm)
  }

  ### sample populations and ploidy
  if (sampleSize == 'equal'){
    
    #create pop_list
    popProb <- rep(1/npop,times=npop)
    pop_list <- rep(1:npop,each=nind/npop)
    
    #incase ind%%npop != 0
    if (nind%%npop != 0){
      pop_list <- c(pop_list,sample(1:npop,nind%%npop,replace = FALSE))
    }
  
    
  }else if (sampleSize == 'unequal'){
    
    #create popProb if not provided
    if (is.null(popProb)){
      popProb <- rep(1/npop,times=npop)
    }
    
    #create pop_list 
    pop_list <- sort(sample(1:npop,nind,replace=TRUE,prob=popProb))
    pops <- unique(pop_list)
    
  }else{stop('not correct sampling scheme option')}
  

  # generate individual genotypes assuming diploid (sample 2)
  sim.geno <- array(0, dim=c(nloci*2, nind)) 
  for (i in (1:nind)){
    sim.geno[,i] <- rbinom(nloci*2, 2, prob=sim.pop[,pop_list[i]])
  }
  
  #filter on minor allele freq (MAF)
  maf_loci <- apply(sim.geno,1,function(g) (1-(length(which(g==0))/length(g))))
  sim.geno <- sim.geno[-which(maf_loci < MAF),]
  
  #remove invariants
  invar_sites <- which(apply(sim.geno,1,function(g) sd(g,na.rm = TRUE)) == 0)
  sim.geno <- sim.geno[-unique(invar_sites),]
  #print(cat((1-(nrow(sim.geno)/(nloci*2)))*100,
            #'% of sites were invariant and removed.\n....A total of ',
            #nrow(sim.geno),' sites were kept for analysis'))
  
  #remove excess loci to make equal to nloci 
  sim.geno <- sim.geno[sample(1:nrow(sim.geno),nloci),]
  
  #create max missing data 
  if(!is.null(maxMissing)){
    missPerc <- rbeta(nind,2,5)*maxMissing
    for (i in 1:nind){
      na_index <- sample(1:nloci,nloci*missPerc[i])
      sim.geno[na_index,i] <- NA
    }
  }
  
  sim_out <- list('geno'=sim.geno,'pop_list'=pop_list,
                  'nloci'=nloci,'nind'=nind,'npop'=npop)
  return(sim_out)
}
```

### PCA function

Makes a PCA with different standardizations.

#### Method options

-   **none:** no center or no standardization\
-   **center:** centering only, no standardization
-   **stand:** center and standardizing\
-   **stand_af:** standardization by rate proportional to expectation of genetic drift (demoninator = sqrt(af\*(1-af))) (Patterson et al. 2006)

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=5,fig.height=5}

## choose colors for PCA 
color_choose <- function(num){
  col10 <- pal_npg()(10)
  col20 <- pal_d3(palette = 'category20')(20)
  if (num <= 10){
    col_out <- col10[1:num]
  }else if (num <= 20){
    col_out <- col20[1:num]
  }else{
    col_out <- hue_pal()(num)
  }
  return(col_out)
}

## PCA_figure (makes a figure on PC1 v PC2) 
PCA_fig <- function(pca_df,pve,title){
  col <- color_choose(length(unique(pca_df$Pop)))
  xlab <- paste("PCA1 (",round(pve[1]*100,2),"%)",sep="")
  ylab <- paste("PCA2 (",round(pve[2]*100,2),"%)",sep="")
  pca_plot <- ggplot(data = pca_df, aes(x=PC1,y=PC2,fill=Pop)) + 
    geom_point(colour='black',pch=21,size = 4) +
    scale_fill_manual(name='Population:',values=col) + 
    xlab(xlab) + ylab(ylab) +
    ggtitle(title) + 
    theme_bw() + 
    theme(#legend.position = 'none',
      axis.text = element_text(size=13), 
      axis.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      plot.title = element_text(size = 16, colour="black",face = "bold"),
      panel.border = element_rect(size = 1.5, colour = "black"),
      legend.text = element_text(size=13),
      legend.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())
  #print(pca_plot)
  return(pca_plot)
}

PCA_ploidy <- function(sim_out,method='none'){
  
  ## initialize variables
  geno <- sim_out$geno
  pop_list <- sim_out$pop_list
  nind <- sim_out$nind
  npop <- sim_out$npop
  
  ## run method and PCA
  if(method == 'none'){
    g_t <- t(geno)
                 
    #fix invariant sites (should not be any)
    g_t[is.na(g_t)] <- 0 
    
    pca_out <- prcomp(g_t,scale. = FALSE,center=FALSE)
    pve <- summary(pca_out)$importance[2,1:5]
    pca_df <- data.frame(Pop=as.character(pop_list),
                         pca_out$x[,1:5])
    title <- 'No center or standardizing'
    
  }else if(method == 'center'){
    g_t <- t(geno)
    
    #center
    g_c <- apply(g_t,2,function(d) (d-mean(d)))
                 
    #fix invariant sites (should not be any)
    g_c[is.na(g_c)] <- 0 
    
    pca_out <- prcomp(g_c,scale. = FALSE,center=FALSE)
    pve <- summary(pca_out)$importance[2,1:5]
    pca_df <- data.frame(Pop=as.character(pop_list),
                         pca_out$x[,1:5])
    title <- 'Centering only'
    
  }else if(method == 'stand'){
    g_t <- t(geno)
    
    #center and standardize
    g_z <- apply(g_t,2,function(d) (d-mean(d))/sd(d))
    
    #fix invariant sites (should not be any)
    g_z[is.na(g_z)] <- 0 
    g_z[is.infinite(g_z)] <- 0 
    
    pca_out <- prcomp(g_z,scale. = FALSE,center=FALSE)
    pve <- summary(pca_out)$importance[2,1:5]
    pca_df <- data.frame(Pop=as.character(pop_list),
                         pca_out$x[,1:5])
    title <- 'Centered and standardized'
    
  }else if(method == 'stand_af'){
    g_t <- t(geno)
    
    #center and standardize by drift
    colmean<-apply(g_t,2,mean,na.rm=TRUE)
    normalize<-matrix(nrow = nrow(g_t),ncol=ncol(g_t))
    af<-colmean/2
    for (m in 1:length(af)){
      nr<-g_t[ ,m]-colmean[m]
      dn<-sqrt(af[m]*(1-af[m]))
      normalize[ ,m]<-nr/dn
    }
    #fix invariant sites (should not be any)
    normalize[is.na(normalize)]<-0
    normalize[is.infinite(normalize)] <- 0 

    pca_out <- prcomp(normalize,scale. = FALSE,center=FALSE)
    pve <- summary(pca_out)$importance[2,1:5]
    
    pca_df <- data.frame(Pop=as.character(pop_list),
                         pca_out$x[,1:5])
    title <- 'Standardization by drift'
    
  }else{
    stop('method not correct')
  }
  pca_plot <- PCA_fig(pca_df,pve,title)
  pca_out <- list("pca_plot"=pca_plot,"pca_df"=pca_df,'pve'=pve)
  return(pca_out)
}

```

### 2 pops, little divergence (Fst = 0.05)

**NOTE:** PVE of the first axis is roughly equal to Fst in a two population model

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=15,fig.height=5}
sim_pop2_fst5 <- sim_mixedploidy(nloci = 5000,nind = 300,npop = 2,
                                 Fst=0.05,sampleSize = 'equal')
```

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=11,fig.height=12}
pca_none2_fst5 <- PCA_ploidy(sim_pop2_fst5,method='none')
pca_center2_fst5 <- PCA_ploidy(sim_pop2_fst5,method='center')
pca_stand2_fst5 <- PCA_ploidy(sim_pop2_fst5,method='stand')
pca_stand_af2_fst5 <- PCA_ploidy(sim_pop2_fst5,method='stand_af')

ggarrange(pca_none2_fst5$pca_plot,pca_center2_fst5$pca_plot,
          pca_stand2_fst5$pca_plot,pca_stand_af2_fst5$pca_plot,
          nrow=3,ncol=2,common.legend = TRUE,align='hv')

```

### 2 pops, high divergence (Fst = 0.2)

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=15,fig.height=5}
sim_pop2_fst2 <- sim_mixedploidy(nloci = 5000,nind = 300,npop = 2,
                                 Fst=0.2,sampleSize = 'equal')
```

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=11,fig.height=12}
pca_none2_fst2 <- PCA_ploidy(sim_pop2_fst2,method='none')
pca_center2_fst2 <- PCA_ploidy(sim_pop2_fst2,method='center')
pca_stand2_fst2 <- PCA_ploidy(sim_pop2_fst2,method='stand')
pca_stand_af2_fst2 <- PCA_ploidy(sim_pop2_fst2,method='stand_af')

ggarrange(pca_none2_fst2$pca_plot,pca_center2_fst2$pca_plot,
          pca_stand2_fst2$pca_plot,pca_stand_af2_fst2$pca_plot,
          nrow=2,ncol=2,common.legend = TRUE,align='hv')

```

### 8 pops, little divergence (Fst = 0.05)

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=15,fig.height=5}
sim_pop8_fst5 <- sim_mixedploidy(nloci = 5000,nind = 300,npop = 8,
                                 Fst=0.05,sampleSize = 'equal')
```

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=11,fig.height=12}
pca_none8_fst5 <- PCA_ploidy(sim_pop8_fst5,method='none')
pca_center8_fst5 <- PCA_ploidy(sim_pop8_fst5,method='center')
pca_stand8_fst5 <- PCA_ploidy(sim_pop8_fst5,method='stand')
pca_stand_af8_fst5 <- PCA_ploidy(sim_pop8_fst5,method='stand_af')

ggarrange(pca_none8_fst5$pca_plot,pca_center8_fst5$pca_plot,
          pca_stand8_fst5$pca_plot,pca_stand_af8_fst5$pca_plot,
          nrow=2,ncol=2,common.legend = TRUE,align='hv')

```

### 8 pops, high divergence (Fst = 0.2)

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=15,fig.height=5}
sim_pop8_fst2 <- sim_mixedploidy(nloci = 5000,nind = 300,npop = 8,
                                 Fst=0.2,sampleSize = 'equal')
```

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=11,fig.height=12}
pca_none8_fst2 <- PCA_ploidy(sim_pop8_fst2,method='none')
pca_center8_fst2 <- PCA_ploidy(sim_pop8_fst2,method='center')
pca_stand8_fst2 <- PCA_ploidy(sim_pop8_fst2,method='stand')
pca_stand_af8_fst2 <- PCA_ploidy(sim_pop8_fst2,method='stand_af')

ggarrange(pca_none8_fst2$pca_plot,pca_center8_fst2$pca_plot,
          pca_stand8_fst2$pca_plot,pca_stand_af8_fst2$pca_plot,
          nrow=2,ncol=2,common.legend = TRUE,align='hv')

```

## Issues of unequal sample size

#### 5 pop, medium divergence (Fst = 0.1), unequal sample sizes.

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=15,fig.height=5}
sim_pop5_equal <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'equal')
sim_pop5_unequal <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'unequal')
sim_pop5_oneLarge <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'unequal',popProb = c(.6,.1,.1,.1,.1))
sim_pop5_twoLarge <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'unequal',popProb = c(.3,.3,.1,.1,.1))
```

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=11,fig.height=12}
pca_equal5 <- PCA_ploidy(sim_pop5_equal,method='stand_af')
pca_equal5_plot <- PCA_fig(pca_equal5$pca_df,pca_equal5$pve,'Equal sample size')
pca_unequal5 <- PCA_ploidy(sim_pop5_unequal,method='stand_af')
pca_unequal5_plot <- PCA_fig(pca_unequal5$pca_df,pca_unequal5$pve,'Unequal sample size')
pca_oneLarge5 <- PCA_ploidy(sim_pop5_oneLarge,method='stand_af')
pca_oneLarge5_plot <- PCA_fig(pca_oneLarge5$pca_df,pca_oneLarge5$pve,'One large population')
pca_twoLarge5 <- PCA_ploidy(sim_pop5_twoLarge,method='stand_af')
pca_twoLarge5_plot <- PCA_fig(pca_twoLarge5$pca_df,pca_twoLarge5$pve,'Two large populations')

ggarrange(pca_equal5_plot,pca_unequal5_plot,
          pca_oneLarge5_plot,pca_twoLarge5_plot,
          nrow=2,ncol=2,common.legend = TRUE,align='hv')

```

## Issues of missing data

#### 5 pop, medium divergence (Fst = 0.1), missing data varies

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=15,fig.height=5}
sim_miss0 <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'equal',maxMissing = NULL)
sim_miss20 <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'equal',maxMissing = .2)
sim_miss50 <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'equal',maxMissing = .5)
sim_miss80 <- sim_mixedploidy(nloci = 5000,nind = 100,npop = 5,
                                 Fst=0.1,sampleSize = 'equal',maxMissing = .8)
```

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=11,fig.height=12}
### make missing PCA function
PCA_miss <- function(sim_out,title){
  
  ## initialize variables
  geno <- sim_out$geno
  pop_list <- sim_out$pop_list
  nind <- sim_out$nind
  npop <- sim_out$npop
  g_t <- t(geno)
  
  ## calculate missing data
  miss <- apply(g_t,1,function(g) length(which(is.na(g)))/ncol(g_t))

  #center and standardize by drift
  colmean<-apply(g_t,2,mean,na.rm=TRUE)
  normalize<-matrix(nrow = nrow(g_t),ncol=ncol(g_t))
  af<-colmean/2
  for (m in 1:length(af)){
    nr<-g_t[ ,m]-colmean[m]
    dn<-sqrt(af[m]*(1-af[m]))
    normalize[ ,m]<-nr/dn
  }
  #fix invariant sites (should not be any)
  normalize[is.na(normalize)]<-0
  normalize[is.infinite(normalize)] <- 0 

  pca_out <- prcomp(normalize,scale. = FALSE,center=FALSE)
  pve <- summary(pca_out)$importance[2,1:5]
    
  pca_df <- data.frame(Pop=as.character(pop_list),
                       miss = miss,pca_out$x[,1:5])
  
  col <- wes_palette("Zissou1", 1000, type = "continuous")
  xlab <- paste("PCA1 (",round(pve[1]*100,2),"%)",sep="")
  ylab <- paste("PCA2 (",round(pve[2]*100,2),"%)",sep="")
  pca_plot <- ggplot(data = pca_df, aes(x=PC1,y=PC2,fill=miss)) + 
    geom_point(colour='black',pch=21,size = 4) +
    scale_fill_gradientn(name='Missing\ndata:',colours=col) + 
    xlab(xlab) + ylab(ylab) +
    ggtitle(title) + 
    theme_bw() + 
    theme(#legend.position = 'none',
      axis.text = element_text(size=13), 
      axis.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      plot.title = element_text(size = 16, colour="black",face = "bold"),
      panel.border = element_rect(size = 1.5, colour = "black"),
      legend.text = element_text(size=13),
      legend.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())
  #print(pca_plot)
  return(pca_plot)
  
}



pca_miss0_plot <- PCA_miss(sim_miss0,title='0% missing data')
pca_miss20_plot <- PCA_miss(sim_miss20,title='20% max missing data')
pca_miss50_plot <- PCA_miss(sim_miss50,title='50% max missing data')
pca_miss80_plot <- PCA_miss(sim_miss80,title='80% max missing data')


ggarrange(pca_miss0_plot,pca_miss20_plot,
          pca_miss50_plot,pca_miss80_plot,
          nrow=2,ncol=2,align='hv')

```

### Color example from Figure 1 in dumb paper

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.width=10,fig.height=5}
pca_color_df <- data.frame(Color=c('red','green','blue','black'),
                           PC1=c(0,-.71,0.71,0),
                           PC2=c(0.82,-.41,-0.41,0),
                           PC3=c(0.14,0.14,0.14,-0.43),
                           OG1=c(1,0,0,0),
                           OG2=c(0,1,0,0),
                           OG3=c(0,0,1,0))
pca_color_df$Color <- factor(pca_color_df$Color,
                             levels=c('red','green','blue','black'))
pve <- c(.44,.44,.11)

col <- c('red','green','blue','black')
pca12_color_plot <- ggplot(data = pca_color_df, aes(x=PC1,y=PC2,fill=Color)) + 
  geom_point(colour='black',pch=21,size = 4) +
  scale_fill_manual(name='Color:',values=col) + 
  xlab(paste("PCA1 (",round(pve[1]*100,2),"%)",sep="")) + 
  ylab(paste("PCA2 (",round(pve[2]*100,2),"%)",sep="")) +
  ggtitle('Color example Fig. 1: PC1 v PC2') + 
  theme_bw() + 
  theme(legend.position = 'none',
      axis.text = element_text(size=13), 
      axis.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      plot.title = element_text(size = 16, colour="black",face = "bold"),
      panel.border = element_rect(size = 1.5, colour = "black"),
      legend.text = element_text(size=13),
      legend.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())
#pca12_color_plot

pca23_color_plot <- ggplot(data = pca_color_df, aes(x=PC2,y=PC3,fill=Color)) + 
  geom_point(colour='black',pch=21,size = 4) +
  scale_fill_manual(name='Color:',values=col) + 
  xlab(paste("PCA2 (",round(pve[2]*100,2),"%)",sep="")) + 
  ylab(paste("PCA3 (",round(pve[3]*100,2),"%)",sep="")) +
  ggtitle('Color example Fig. 1: PC2 v PC3') + 
  theme_bw() + 
  theme(legend.position = 'none',
      axis.text = element_text(size=13), 
      axis.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      plot.title = element_text(size = 16, colour="black",face = "bold"),
      panel.border = element_rect(size = 1.5, colour = "black"),
      legend.text = element_text(size=13),
      legend.title = element_text(size = 16, colour="black",face = "bold",vjust = 1),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())

pca12_color_plot + pca23_color_plot


#### Look at distances of PCA and OG matrix 

dist_color <- dist(pca_color_df[,5:7])
dist_color

dist_color_pca <- dist(pca_color_df[,2:4])
dist_color_pca

cor(as.vector(dist_color),as.vector(dist_color_pca))

plot(as.vector(dist_color),as.vector(dist_color_pca))

```
