---
output:
  html_document: default
  pdf_document: default
---


```{r, echo=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```
---
title: "Function to extract entropy output and make figure"
author: "Trevor Faske"
output: pdf_document
---

This Rmarkdown provides a function and example code extract entropy output for any number of k and make structure-like figures.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Desktop/src/R/Rmarkdown/entropy_out/')
```

### Loading packages

```{r eval=TRUE, eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE}
library(tidyverse) #i hate it too 
library(ggsci) #color_choose
library(patchwork) #multipanel figures
```

# entropy_out function

#### inputs:    
- **q**: q data.frame (nrow = nind*k) 
  - entropy command ex.: `estpost *k2*.hdf5 -p q -s 0 -o q2.txt`\  
- **MCMC**: MCMC data.frame  
  - entropy command ex.: `estpost *k2*.hdf5 -p q -s 4 -o MCMC_k2.txt`\    
- **k**: K value\   
- **indv_list**: list of individual names in same order as vcf file\    
- **pop_id**: (optional) data.frame containing information for each individual in same order as vcf file to combine anc data to and potentially color,sort figure by\  
- **pop_id_indv**: (required if pop_id present) column name which is equal to indv_list\    
- **figure:** (TRUE/FALSE, default = FALSE) TRUE if you want to make structure like figure\     
- **sort_by:** (optional) column in pop_id to sort figure by (e.g. PCA1, Lat)\  
- **group:** (optional) column in pop_id to group structure figure by (e.g. Population, Species)\  
- **colors:** (optional) list of colors for structure figure equal to k. Default: ggsci npg for k <= 10, ggsci d3(palette='category20') for 10 < k <= 20, and ggplot default for k > 20  

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE}

######################################## entropy_out function ########################################

entropy_out <- function(q,MCMC,k,indv_list,pop_id=NULL,pop_id_indv=NULL,
                        figure=TRUE,sort_by=NULL,group=NULL,colors=NULL){
  
  # add indv_list and MCMC summary stats to q
  q <- q[,1:2]
  names(q)[2] <- 'q'
  q <- left_join(q,MCMC)
  q[,1] <- as.character(indv_list)
  q$Anc <- rep(paste0('A',1:k),each=length(indv_list))
  q$ESS <- rep(paste0('ESS',1:k),each=length(indv_list))
  q$Rhat <- rep(paste0('Rhat',1:k),each=length(indv_list))
  
  # join file with pop_id if present
  if (is.null(pop_id)){
    pop_id_indv <- 'All'
    names(q)[1] <- pop_id_indv
    k_out <- q
  }else{
    if(is.null(pop_id_indv)){
      stop('Must provide column name in pop_id_indv')
    }
    names(q)[1] <- pop_id_indv
    k_out <- left_join(pop_id,q)
  }
  
  # make k_out wide (sure there is a better way to do this)
   spec1 <- k_out %>%
      build_wider_spec(names_from = c(Anc,ESS,Rhat), 
                values_from = c(q,effectiveSampleSize,potentialScaleReductionFactor))
   spec1$.name <- c(unique(k_out$Anc),unique(k_out$ESS),unique(k_out$Rhat))
   k_out_wide <- k_out %>% 
     pivot_wider_spec(spec1)

   # if need to make figure
   if (figure){
     #sort by ancestry (q) so bars make sense
     k_out_wide_sort <-  k_out_wide[order(k_out_wide[unique(k_out$Anc)[-length(unique(k_out$Anc))]],
                                         decreasing=TRUE),]
     k_out_long <- as.data.frame(k_out)
     k_out_long[[pop_id_indv]] <-  factor(k_out_long[[pop_id_indv]],
                                        levels=k_out_wide_sort[[pop_id_indv]])

     #group and sort_by
     if (!(is.null(group)) & !(is.null(sort_by))){ #both
        #print('both contain something')
        group_order <- as.character(unique(k_out_long[[group]][order(k_out_long[[sort_by]])]))
        k_out_long[[group]] <- factor(k_out_long[[group]],levels = group_order)
     }else if(!(is.null(sort_by))){ #sort_by only
       #print('sort_by contains something')
       k_out_long <- k_out_long[order(-1*k_out_long[[sort_by]]),]
       k_out_long[[pop_id_indv]] <-  factor(k_out_long[[pop_id_indv]],
                                        levels=k_out_wide_sort[[pop_id_indv]])
     }else if(!(is.null(group))){# group only
       #print('group contains something')
     }else{#neither
       
     }
     #run figure function 
     ent_fig <- entropy_fig(k_out_long,pop_id_indv,group,colors)
   }else{
     ent_fig=NULL
   }
   
   return(list(ent_out = k_out_wide,ent_fig=ent_fig))
}

######################################## entropy_figure function ########################################

entropy_fig <- function(k_out_long,pop_id_indv,group=NULL,colors=NULL){
  
  # specify colors
  if (is.null(colors)){
    col <- color_choose(length(unique(k_out_long$Anc)))
  }else{
    col <- colors
  }

  # Make figure 
  ent_fig <- ggplot(k_out_long, aes(x=k_out_long[[pop_id_indv]], y=q, fill=Anc)) +
      geom_bar(stat="identity", position="fill") +
      scale_x_discrete(expand=c(0, 0)) +
      scale_y_continuous(expand=c(0, 0)) +
      scale_fill_manual(values=col) +
      guides(fill=guide_legend(override.aes=list(colour=NULL))) +
      theme_bw()+
      theme(panel.grid=element_blank(),
            panel.background=element_rect(fill=NA, colour="black"),
            panel.spacing=unit(0.15, "lines"),
            axis.title=element_blank(),
            axis.text.x=element_blank(),
            axis.text = element_text(size=13),
            axis.ticks.x=element_blank(),
            strip.background=element_blank(),
            #rect = element_rect(fill = "transparent"),
            strip.text=element_text(size=13,colour="black",face='bold'),
            legend.position = 'none',
            legend.key=element_rect(colour="black"))
    #print(ent_fig)
  
  #add group if exists
  if(!is.null(group)){ 
    ent_fig <- ent_fig + 
      facet_grid(~k_out_long[[group]], switch = "x", scales = "free", space = "free")
  }
  return(ent_fig)
}

######################################## color_choose function ########################################

color_choose <- function(num){
  col10 <- pal_npg()(10)
  col20 <- pal_d3(palette = 'category20')(20)
  if (num <= 10){
    col_out <- col10[1:num]
  }else if (num <= 20){
    col_out <- col20[1:num]
  }else{
    col_out <- hue_pal()(num)
  }
  return(col_out)
}
```

## Examples  

### ERNA (*Ericameria nauseosa*, rubber rabbitbrush)  

#### Features:    
- 585 individuals, 75 populations  
- extremely broad sampling  
- sort_by: PC1  
- group: NULL    
- colors: given  
- k: only one  

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.height=3,fig.width=12}

# Set variables
q <- read.csv("data/ERNA/q2.txt", header=TRUE)
MCMC <- read.csv('data/ERNA/MCMC_k2.txt')
k <- 2
pop_id <- read.csv('data/ERNA/pca_df.csv')
indv_list <- as.character(pop_id$All)
pop_id_indv <- 'All'
figure <- TRUE
sort_by <- 'PC1'
group <- NULL
colors <- c('grey40','seagreen')

# Run function
ERNA_k2 <- entropy_out(q,MCMC,k,indv_list,pop_id,pop_id_indv,
                        figure,sort_by,group,colors)

# data frame 
write.csv(ERNA_k2$ent_out,'data/ERNA/k2_out.csv',row.names=FALSE)

# figure
ERNA_k2_fig <-ERNA_k2$ent_fig
ERNA_k2_fig

ggsave('data/ERNA/k2_figure.pdf',height=3,width=10)

```

### CHDO (*Chaenactis douglasii*, dusty maiden):   

#### Features:    
- 272 individuals, 24 populations, 3 ploidy levels
- extremely broad sampling  
- sort_by: PC1  
- group: population and ploidy   
- colors: NULL  
- k: three (k2,k3,k4) 

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.height=9,fig.width=12}

# Set variables
q2 <- read.csv("data/CHDO/q2.txt", header=TRUE)
MCMC2 <- read.csv('data/CHDO/MCMC_k2.txt')
q3 <- read.csv("data/CHDO/q3.txt", header=TRUE)
MCMC3 <- read.csv('data/CHDO/MCMC_k3.txt')
q4 <- read.csv("data/CHDO/q4.txt", header=TRUE)
MCMC4 <- read.csv('data/CHDO/MCMC_k4.txt')

pop_id <- read.csv('data/CHDO/pca_df.csv')
indv_list <- as.character(pop_id$All)
pop_id_indv <- 'All'
figure <- TRUE
sort_by <- 'PC1'
group <- 'PopPloidy'
colors <- NULL

#### k2 ####
# Run function
CHDO_k2 <- entropy_out(q=q2,MCMC=MCMC2,k=2,indv_list,pop_id,pop_id_indv,
                        figure,sort_by,group,colors)

# data frame 
write.csv(CHDO_k2$ent_out,'data/CHDO/k2_out.csv',row.names=FALSE)

# figure
CHDO_k2_fig <-CHDO_k2$ent_fig
#CHDO_k2_fig

#ggsave('data/CHDO/k2_figure.pdf',height=3,width=12)

#### k3 ####
# Run function
CHDO_k3 <- entropy_out(q=q3,MCMC=MCMC3,k=3,indv_list,pop_id,pop_id_indv,
                        figure,sort_by,group,colors)

# data frame 
write.csv(CHDO_k3$ent_out,'data/CHDO/k3_out.csv',row.names=FALSE)

# figure
CHDO_k3_fig <-CHDO_k3$ent_fig
#CHDO_k3_fig

#ggsave('data/CHDO/k3_figure.pdf',height=3,width=12)

#### k4 ####
# Run function
CHDO_k4 <- entropy_out(q=q4,MCMC=MCMC4,k=4,indv_list,pop_id,pop_id_indv,
                        figure,sort_by,group,colors)

# data frame 
write.csv(CHDO_k4$ent_out,'data/CHDO/k4_out.csv',row.names=FALSE)

# figure
CHDO_k4_fig <-CHDO_k4$ent_fig
#CHDO_k4_fig

#ggsave('data/CHDO/k4_figure.pdf',height=3,width=14)

#### Put all figures together

CHDO_k2_fig / CHDO_k3_fig / CHDO_k4_fig

ggsave('data/CHDO/k2-4_figure.pdf',height=9,width=12)

```

### PIRE: sort_by, group, no colors, multiple ks 

#### Features:    
- 149 individuals, 24 populations, 3 ploidy levels  
- extremely fine-scale sampling  
- sort_by: Latitude  
- group: Trail ID   
- colors: NULL
- k: very high (k10,k11) 

```{r eval=TRUE, echo=TRUE,warning=FALSE,error=FALSE,message=FALSE,fig.height=6,fig.width=12}

# Set variables
q10 <- read.csv("data/PIRE/q10.txt", header=TRUE)
MCMC10 <- read.csv('data/PIRE/MCMC_k10.txt')
q11 <- read.csv("data/PIRE/q11.txt", header=TRUE)
MCMC11 <- read.csv('data/PIRE/MCMC_k11.txt')

pop_id <- read.csv('data/PIRE/pca_df.csv')
indv_list <- as.character(pop_id$All)
pop_id_indv <- 'All'
figure <- TRUE
sort_by <- 'Lat'
group <- 'Trail'
colors <- NULL

#### k10 ####
# Run function
PIRE_k10 <- entropy_out(q=q10,MCMC=MCMC10,k=10,indv_list,pop_id,pop_id_indv,
                        figure,sort_by,group,colors)

# data frame 
write.csv(PIRE_k10$ent_out,'data/PIRE/k10_out.csv',row.names=FALSE)

# figure
PIRE_k10_fig <-PIRE_k10$ent_fig
#PIRE_k10_fig

#ggsave('data/PIRE/k10_figure.pdf',height=3,width=12)

#### k11 ####
# Run function
PIRE_k11 <- entropy_out(q=q11,MCMC=MCMC11,k=11,indv_list,pop_id,pop_id_indv,
                        figure,sort_by,group,colors)

# data frame 
write.csv(PIRE_k11$ent_out,'data/PIRE/k11_out.csv',row.names=FALSE)

# figure
PIRE_k11_fig <-PIRE_k11$ent_fig
#PIRE_k11_fig

#ggsave('data/PIRE/k11_figure.pdf',height=11,width=12)

#### Put all figures together

PIRE_k10_fig / PIRE_k11_fig

ggsave('data/PIRE/k10-11_figure.pdf',height=6,width=12)

```
